---
title: "ESM 244 Lab 2 - Time series"
author: "M. D. Mohr"
format: 
  html: 
    code-fold: show
    toc: true
    number-sections: true
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
library(here)
library(tsibble)
library(feasts)
library(fable)
```

# Part 1: Time series with Toolik Lake data

## Always look at your data

### Read in the data

```{r}
toolik_df <- read_csv(here('data/toolik_daily.csv'))

### convert data frame to time series
toolik_ts <- toolik_df %>% 
  mutate(date = lubridate::mdy(date)) %>% 
  as_tsibble(key = NULL, ### if we had multiple sites, key by site
             index = date) ### this is our time series variable
  
ggplot(toolik_ts, aes(x=date, y = daily_air_temp)) +
  geom_line() +
  labs(x = 'Date', y = 'Mean daily air temp (Celsius)\n at Toolik Station')
```

## Use the filter_index() function to filter by date/time!

```{r}
### Filter from Dec 2010 to Jan 2011 
toolik_ts %>% 
  filter_index("2018-01-01" ~ .)
```

## Use index_by() to aggregate time series by increments
```{r}
toolik_month <- toolik_ts %>% 
  index_by(yr_mo = ~yearmonth(.)) %>%
  summarize(monthly_mean_temp = mean(daily_air_temp, na.rm = TRUE)) %>%
  ungroup()
```

Now let's plot that! 

```{r}
ggplot(data = toolik_month, aes(x = yr_mo, y = monthly_mean_temp)) + 
  geom_line()

ggplot(toolik_month, aes (x = year(yr_mo), y = monthly_mean_temp)) +
  geom_line()+
  facet_wrap(~month(yr_mo, label = TRUE)) + 
  labs( x = 'Year', y = 'Annual mean air temp (Celsius)', 
        title = 'Toolik Station mean annual air temperature',
        subtitle = '1988-2023',
        caption = '<put citation here>')
```

# Part 2: Time series wrangling an forecasting!

Energy usage by sector, in trillions of BTUs.

```{r}
energy_df <- read_csv(here('data', 'energy.csv'))
```

### Analysis goals:

* Examine patterns and trends in residential energy consumption over time
* Predict what residential energy use patterns will look like over the next 5 years

### Pseudocode

My responses: 
* create a new ts data frame
* modify (mutate) the date column to be in the standard 
* filter sector column to have residential observation only 
* plot the time series data to view the trends using the modified data frame 
* no clue how to do the forecasting part 

In class responses: 
* convert year-moth to a date
* average the energy consumption by year, so looking year-to-year
* develop an exploratory ggplot for trends, seasonality, etc
* filter the dataset to look at residential 
* do forecasting once we have looked at the trends
* identify trend, seasonality (before forecasting)
* convert to time series using as_tsibble 
* Define the alpha value (from the exponential smoothing) ad select a suitable model
* decompose the data to look for trends and seasonality

```{r}
energy_ts <- energy_df %>% 
  mutate(date = tsibble::yearmonth(yrmonth)) %>% 
  as_tsibble(index = date, 
             key = sector)
```

```{r}
ggplot(data = energy_ts, aes(x = date, y = energy_total, color = sector)) + 
  geom_line() +
  labs(y = "Energy consumption by sector \n (Trillion BTUS)") +
  facet_wrap(~ sector, ncol = 1)
```

* Residential looks similar to commercial, with an upward trend at least for the first part, maybe leveling off or decreasing in recent years
* Seasonality - summer and winter peaks, summer peaks seem to be getting larger over time 

### Season plot

```{r}
energy_ts %>% 
  filter(sector == 'residential') %>% 
  gg_season(y = energy_total, pal = hcl.colors(n=9)) + 
  theme_light() +
  labs( x = "Month", y = "Residential energy consumption (trillion BTU)")
```

### Subseries Plot
```{r}
energy_ts %>% 
  gg_subseries(y = energy_total)
```


## Decomposition

```{r}
### Find the STL decomposition (L = LOESS) locally estimated scatterplot smoothing

dcmp <- energy_ts %>% 
  filter(sector == 'residential') %>%
  model(feasts::STL(energy_total ~ season(period = '1 year') +
                      trend(window = 49)))

components(dcmp) %>% 
  autoplot()

```


### Autocorrelation function
```{r}
energy_ts %>%
  filter(sector == 'residential') %>%
  ACF(energy_total) %>% 
  autoplot()
```

## Forecasting by Holds Winters exponential smoothing

Specify for each component, whether none ("N"), additice ("A"), or multiplicative ("M") 
```{r}
### Create a model
energy_fit <- energy_ts %>% 
  filter(sector == 'residential') %>% 
  filter_index('2000-01' ~.) %>%
  model(ets = ETS(energy_total ~ season(method = "M") + trend(method = "A")))

energy_forecast <- energy_fit %>% 
  forecast(h = "5 years")
energy_forecast %>% 
  autoplot(energy_ts)
```

```{r}
energy_predicted <- energy_fit %>% 
  broom::augment()

ggplot(energy_predicted) +
  geom_line(aes(x = date, y = energy_total)) +
  geom_line(aes(x = date, y = .fitted), color = 'red', alpha = .7)

ggplot(energy_predicted, aes(x = .resid)) + 
  geom_histogram()
```

